### 2022-02-25 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_networkary_classifiers@503c15f72088d6e5443c3b54bbd5062fe51628d3 🚀

The accuracy paradox is the basis of a number of problems in statistical
education, and lies in the fact that, when the desired class is rare, a
model that gets less and less performant will become more and more
accurate and useful, simply by (i) underpredicting true positive cases
and (ii) over-predicting false negatives. In other words, accuracy,
defined as the proportion of predictions that are correct, is often
useless as a measure of how predictive a model is. This is particularly
true in ecological networks; the desired class (presence of an
interaction between two species) is the one we care most about, and by
far the least commmon. Herein lies the core challenge of predicting
species interactions: the extreme imbalance between classes makes the
training of predictive models difficult, and their validation even more
so as we do not reliably know which negatives are true. The connectance
of empirical networks is usually well under 20\%, with larger networks
having a lower connectance (MacDonald et al., 2020). Recent
contributions (Becker et al., 2021; Strydom et al., 2021) highlight that
predictive models of interactions can likely be improved by adding
information (in the form of, \emph{e.g.} traits), but that we do not
have robust guidelines as to how the predictive ability of these models
should be evaluated, nor about how the models should be trained. Here,
by relying on simple derivations and a series of simulations, we
formulate a number of such guidelines, specifically for the case of
binary classifiers derived from thresholded values.

@@

Informedness (Youden, 1950) (also known as bookmaker informedness or the
True Skill Statistic) is \(\text{TPR}+\text{TNR}-1\), where
\(\text{TPR}= tp/(tp+fn)\) and \(\text{TNR} = tn/(tn+fp)\); informedness
can be used to find the optimal cutpoint in thresholding analyses
(Schisterman et al., 2005). The MCC is defined as

@@

The same approach is used to evaluate \emph{e.g.} species distribution
models (SDMs). Indeed, the training and evaluation of SDMs as binary
classifiers suffers from the same issue of low prevalence; this is not
surprising that the two fields (SDMs and network predictions) would
share methods and their attached conceptual issues, as they suffer from
data limitations, class imbalance, and the conversion of quantitative
prediction into a binary classification. In previous work, Allouche et
al. (2006) suggested that \(\kappa\) was a better test of model
performance than the True Skill Statistic (TSS; which we refer to as
Youden's informedness); these conclusions were later criticized by
Somodi et al. (2017), who emphasized that informedness' relationship to
prevalence depends on assumptions about bias in the model, and therefore
recommend the use of \(\kappa\) as a validation of classification
performance. Although this work offers recommendations about the
comparison of models, it doesn't establishes baselines or good practices
for training on imbalanced ecological data. Within the context of
networks, there are three specific issues that need to be adressed.
First, what values of performance measures are we expecting for a
classifier that has poor performance? This is particularly important as
it can evaluate whether low prevalence can lull us into a false sense of
predictive accuracy. Second, independently of the question of model
evaluation, is low prevalence an issue for \emph{training}, and can we
remedy it? Finally, because the low amount of data on interaction makes
a lot of imbalance correction methods (see \emph{e.g.} Branco et al.,
2015) hard to apply, which indicators can be optimized with the least
amount of positive interaction data?

In addition to the literature on SDMs, most of the research on machine
learning application to life sciences is focused on genomics (which has
very specific challenges, see a recent discussion by Whalen et al.,
2021); this sub-field has generated largely different recommendations.
Chicco \& Jurman (2020) suggest using Matthews correlation coefficient
(MCC) over \(F_1\), as a protection against over-inflation of predicted
results; Delgado \& Tibau (2019) advocate against the use of Cohen's
\(\kappa\), again in favor of MCC, as the relative nature of \(\kappa\)
means that a worse classifier can be picked over a better one;
similarly, Boughorbel et al. (2017) recommend MCC over other measures of
performance for imbalanced data, as it has more desirable statistical
properties. More recently, Chicco et al. (2021) temper the apparent
supremacy of the MCC, by suggesting it should be replaced by Youden's
informedness (also known as \(J\), bookmaker's accuracy, and the
True-Skill Statistic) when the imbalance in the dataset may not be
representative (Jordano, 2016a, which is the case as networks are
under-sampled; 2016b), when classifiers need to be compared across
different datasets (for example when predicting a system in space, where
undersampling varies locally; McLeod et al., 2021), and when comparing
the results to a no-skill (baseline) classifier is important. As these
conditions are likely to be met with network data, there is a need to
evaluate which measures of classification accuracy respond in a
desirable way.

@@

In order to examine how MCC, \(F_1\), \(\kappa\), and informedness
change w.r.t. the imbalance, skill, and bias, we performed a grid
exploration of the values of \(\text{logit}(s)\) and \(\text{logit}(b)\)
linearly from \(-10\) to \(10\), of \(\rho\) linearly in \(]0, 0.5]\),
which is within the range of usually observed connectance values for
empirical food webs. Note that at this point, there is no food web model
to speak of; rather, the confusion matrix we discuss can be obtained for
any classification task. Based on the previous discussion, the desirable
properties for a measure of classifier success should be: an increase
with classifier skill, especially at low bias; a hump-shaped response to
bias, especially at high skill, and ideally center around
\(\text{logit}(b)=0\); an increase with prevalence up until
equiprevalence is reached.

@@

In the following section, we will generate random bipartite networks
(this works without loss of generality on unipartite networks), and
train four binary classifiers (as well as an ensemble model using the
sum of ranged outputs from the component models) on 30\% of the
interaction data. Networks are generated by picking a random
infectiousness trait \(v_i\) for 100 species (from a \(B(6,8)\)
distribution), and a resistance trait \(h_j\) for 100 species (from a
\(B(2,8)\) distribution). There is an interaction between \(i\) and
\(j\) when \(v_i-\xi/2 \le h_j \le v_i+\xi/2\), where \(\xi\) is a
constant regulating the connectance of the network (there is an almost
1:1 relationship between \(\xi\) and connectance), and varies uniformly
in \([0.05, 0.35]\). This model gives fully interval networks that are
close analogues to the bacteria--phage model of Weitz et al. (2005),
with both a modular structure and a non-uniform degree distribution.
This model is easy to learn: when trained with features
\([v_i, h_j, \text{abs}(v_i, h_j)] ^T\) to predict the interactions
between \(i\) and \(j\), all four models presented below were able to
reach almost perfect predictions all the time (data not presented here)
-- this is in part because the rule is fixed for all interactions. In
order to make the problem more difficult to solve, we use \([v_i, h_j]\)
as a feature vector (\emph{i.e.} the traits on which the models are
trained), and therefore the models will have to uncover that the rule
for interaction is \(\text{abs}(v_i, h_j) \le \xi\).

@@

Finally, it is noteworthy that the ensemble model was systematically
better than the component models; even when the models were individually
far form perfect, the ensemble was able to leverage the different biases
expressed by the models to make an overall more accurate prediction. We
do not expect that ensembles will \emph{always} be better than single
models. In a recent multi-model comparison, Becker et al. (2021) found
that the ensemble was \emph{not} the best model. There is no general
conclusion to draw from this besides reinforcing the need to be
pragmatic about which models should be included in the ensemble, or
whether to use an ensemble at all. In a sense, the surprising peformance
of the ensemble model should form the basis of the last recommendation:
optimal training set bias and its interaction with connectance and
binary classifier is, in a sense, an hyperparameter that should be
assessed. The distribution of results in fig.~\ref{fig:optimbias} and
fig.~\ref{fig:optimperf} show that there are variations around the
trend; furthermore, networks with different structures than the one we
simulated here may respond in different ways.

@@

\leavevmode\hypertarget{ref-Becker2021OptPre}{}%
Becker, D., Albery, G. F., Sjodin, A. R., Poisot, T., Bergner, L.,
Dallas, T., Eskew, E. A., Farrell, M. J., Guth, S., Han, B. A., Simmons,
N. B., Stock, M., Teeling, E. C., \& Carlson, C. J. (2021). Optimizing
predictive models to prioritize viral discovery in zoonotic reservoirs.
\emph{bioRxiv}, 2020.05.22.111344.
\url{https://doi.org/10.1101/2020.05.22.111344}

@@

\leavevmode\hypertarget{ref-Blaom2020FleMod}{}%
Blaom, A. D., \& Vollmer, S. J. (2020, December 31). \emph{Flexible
model composition in machine learning and its implementation in MLJ}.
\url{http://arxiv.org/abs/2012.15505}

@@

\leavevmode\hypertarget{ref-Branco2015SurPre}{}%
Branco, P., Torgo, L., \& Ribeiro, R. (2015, May 13). \emph{A Survey of
Predictive Modelling under Imbalanced Distributions}.
\url{http://arxiv.org/abs/1505.01658}

@@

\leavevmode\hypertarget{ref-Jeni2013FacImb}{}%
Jeni, L. A., Cohn, J. F., \& De La Torre, F. (2013). Facing Imbalanced
Data--Recommendations for the Use of Performance Metrics. \emph{2013
Humaine Association Conference on Affective Computing and Intelligent
Interaction}, 245--251. \url{https://doi.org/10.1109/ACII.2013.47}

@@

\leavevmode\hypertarget{ref-Poisot2021ImpMam}{}%
Poisot, T., Ouellet, M.-A., Mollentze, N., Farrell, M. J., Becker, D.
J., Albery, G. F., Gibb, R. J., Seifert, S. N., \& Carlson, C. J. (2021,
May 31). \emph{Imputing the mammalian virome with linear filtering and
singular value decomposition}. \url{http://arxiv.org/abs/2105.14973}

@@

\leavevmode\hypertarget{ref-Somodi2017PreDep}{}%
Somodi, I., Lepesi, N., \& Botta‐Dukát, Z. (2017). Prevalence dependence
in model goodness measures with special emphasis on true skill
statistics. \emph{Ecology and Evolution}, \emph{7}(3), 863--872.
\url{https://doi.org/10.1002/ece3.2654}

>>>>

The accuracy paradox is the basis of a number of problems in statistical
education, and lies in the fact that, when the desired class is rare, a
model that gets less and less performant will become more and more
accurate and useful, simply by (i) underpredicting true positive cases
and (ii) over-predicting false negatives. In other words, accuracy,
defined as the proportion of predictions that are correct, is often
useless as a measure of how predictive a model is. This is particularly
true in ecological networks; the desired class (presence of an
interaction between two species) is the one we care most about, and by
far the least commmon. Herein lies the core challenge of predicting
species interactions: the extreme imbalance between classes makes the
training of predictive models difficult, and their validation even more
so as we do not reliably know which negatives are true. The connectance
(the proportion of realized interactions, usually the number of
interactions divided by the number of species pairs) of empirical
networks is usually well under 20\%, with larger networks having a lower
connectance (MacDonald et al., 2020). Recent contributions (Strydom et
al., 2021; \textbf{Becker2021OptPre?}) highlight that predictive models
of interactions can likely be improved by adding information (in the
form of, \emph{e.g.} traits), but that we do not have robust guidelines
as to how the predictive ability of these models should be evaluated,
nor about how the models should be trained. Here, by relying on simple
derivations and a series of simulations, we formulate a number of such
guidelines, specifically for the case of binary classifiers derived from
thresholded values.

@@

Informedness (Youden, 1950) (also known as bookmaker informedness or the
True Skill Statistic) is \(\text{TPR}+\text{TNR}-1\), where
\(\text{TPR}= tp/(tp+fn)\) and \(\text{TNR} = tn/(tn+fp)\); informedness
can be used to find the optimal cutpoint in thresholding analyses
(Schisterman et al., 2005). The formula for informedness is

\[\frac{tp}{tp+fn}+\frac{tn}{tn+fp}-1\,.\]

The MCC is defined as

@@

The same approach is used to evaluate \emph{e.g.} species distribution
models (SDMs). Indeed, the training and evaluation of SDMs as binary
classifiers suffers from the same issue of low prevalence; this is not
surprising that the two fields (SDMs and network predictions) would
share methods and their attached conceptual issues, as they suffer from
data limitations, class imbalance, and the conversion of quantitative
prediction into a binary classification. In previous work, Allouche et
al. (2006) suggested that \(\kappa\) was a better test of model
performance than the True Skill Statistic (TSS; which we refer to as
Youden's informedness); these conclusions were later criticized by
Somodi et al. (2017), who emphasized that informedness' relationship to
prevalence depends on assumptions about bias in the model, and therefore
recommend the use of \(\kappa\) as a validation of classification
performance. Although this work offers recommendations about the
comparison of models, it doesn't establishes baselines or good practices
for training on imbalanced ecological data. Steen et al. (2021) show
that, when applying spatial thinning (a process that has no analogue in
networks), the best approach to train ML-based SDMs varies according to
the balancing of the dataset, and the evaluation measures used. This
suggests that there is no single ``recipe'' that is guaranteed to give
the best model. Within the context of networks, there are three specific
issues that need to be adressed. First, what values of performance
measures are we expecting for a classifier that has poor performance?
This is particularly important as it can evaluate whether low prevalence
can lull us into a false sense of predictive accuracy. Second,
independently of the question of model evaluation, is low prevalence an
issue for \emph{training}, and can we remedy it? Finally, because the
low amount of data on interaction makes a lot of imbalance correction
methods (see \emph{e.g.} Branco et al., 2015) hard to apply, which
indicators can be optimized with the least amount of positive
interaction data?

In addition to the literature on SDMs, most of the research on machine
learning application to life sciences is focused on genomics (which has
very specific challenges, see a recent discussion by Whalen et al.,
2021); this sub-field has generated largely different recommendations.
Chicco \& Jurman (2020) suggest using Matthews correlation coefficient
(MCC) over \(F_1\), as a protection against over-inflation of predicted
results; Delgado \& Tibau (2019) advocate against the use of Cohen's
\(\kappa\), again in favor of MCC, as the relative nature of \(\kappa\)
means that a worse classifier can be picked over a better one;
similarly, Boughorbel et al. (2017) recommend MCC over other measures of
performance for imbalanced data, as it has more desirable statistical
properties. More recently, Chicco et al. (2021) temper the apparent
supremacy of the MCC, by suggesting it should be replaced by Youden's
informedness (also known as \(J\), bookmaker's accuracy, and the
True-Skill Statistic) when the imbalance in the dataset may not be
representative, which is the case as species interaction networks are
often under-sampled (Jordano, 2016b, 2016a), when classifiers need to be
compared across different datasets (for example when predicting a system
in space, where undersampling varies locally; McLeod et al., 2021), and
when comparing the results to a no-skill (baseline) classifier is
important. As these conditions are likely to be met with network data,
there is a need to evaluate which measures of classification accuracy
respond in a desirable way.

@@

In order to examine how MCC, \(F_1\), \(\kappa\), and informedness
change w.r.t. the imbalance, skill, and bias, we performed a grid
exploration of the values of \(\text{logit}(s)\) and \(\text{logit}(b)\)
linearly from \(-10\) to \(10\); \(\text{logit}(x) = -10\) means that
\(x\) is essentially 0, and \(\text{logit}(x) = 10\) means it is
essentially 1 -- this choice was motivated by the fact that most
responses are non-linear with regards to bias and skill. The values or
\(\rho\) were taken linearly in \(]0, 0.5]\), which is within the range
of usually observed connectance values for empirical species interaction
networks. Note that at this point, there is no food web model to speak
of; rather, the confusion matrix we discuss can be obtained for any
classification task. Based on the previous discussion, the desirable
properties for a measure of classifier success should be: an increase
with classifier skill, especially at low bias; a hump-shaped response to
bias, especially at high skill, and ideally center around
\(\text{logit}(b)=0\); an increase with prevalence up until
equiprevalence is reached.

@@

In the following section, we will generate random bipartite networks
(this works without loss of generality on unipartite networks), and
train four binary classifiers (as well as an ensemble model using the
sum of ranged outputs from the component models) on 30\% of the
interaction data. In practice, testing usually uses 70\% of the total
data; for ecological networks, where interactions are sparse \emph{and}
the number of species is low, this may not be the best solution, as the
testing set becomes constrained not by the \emph{proportion} of
interactions, but by their \emph{number}. Preliminary experiments using
different splits revealed no qualitative change in the results. Networks
are generated by picking a random infectiousness trait \(v_i\) for 100
species (from a beta distribution \(B(\alpha=6,\beta=8)\) distribution),
and a resistance trait \(h_j\) for 100 species (from
\(B(\alpha=2,\beta=8)\) distribution). There is an interaction between
\(i\) and \(j\) when \(v_i-\xi/2 \le h_j \le v_i+\xi/2\), where \(\xi\)
is a constant regulating the connectance of the network (there is an
almost 1:1 relationship between \(\xi\) and connectance), and varies
uniformly in \([0.05, 0.35]\). This model gives fully interval networks
that are close analogues to the bacteria--phage model of Weitz et al.
(2005), with both a modular structure and a non-uniform degree
distribution. This dataset is easy for almost any algorithm to learn:
when trained with features \([v_i, h_j, \text{abs}(v_i, h_j)] ^T\) to
predict the interactions between \(i\) and \(j\), all four models
presented below were able to reach almost perfect predictions all the
time (data not presented here) -- this is in part because the rule
(there is maximum value of the distance between traits for which there
is an interaction) is fixed for all interactions. In order to make the
problem more difficult to solve, we use \([v_i, h_j]\) as a feature
vector (\emph{i.e.} the traits on which the models are trained), and
therefore the models will have to uncover that the rule for interaction
is \(\text{abs}(v_i, h_j) \le \xi\). The models therefore all have the
form

\[
\begin{bmatrix}
           x_{1} \\
           x_{2} \\
           \vdots \\
           x_{m}
         \end{bmatrix}
\]

@@

Finally, it is noteworthy that the ensemble model was systematically
better than the component models; even when the models were individually
far form perfect, the ensemble was able to leverage the different biases
expressed by the models to make an overall more accurate prediction. We
do not expect that ensembles will \emph{always} be better than single
models. In a recent multi-model comparison, (\textbf{Becker2021OptPre?})
found that the ensemble was \emph{not} the best model. There is no
general conclusion to draw from this besides reinforcing the need to be
pragmatic about which models should be included in the ensemble, or
whether to use an ensemble at all. In a sense, the surprising peformance
of the ensemble model should form the basis of the last recommendation:
optimal training set bias and its interaction with connectance and
binary classifier is, in a sense, an hyperparameter that should be
assessed. The distribution of results in fig.~\ref{fig:optimbias} and
fig.~\ref{fig:optimperf} show that there are variations around the
trend; furthermore, networks with different structures than the one we
simulated here may respond in different ways.

@@

@@

\leavevmode\hypertarget{ref-Blaom2020FleMod}{}%
Blaom, A. D., \& Vollmer, S. J. (2020). Flexible model composition in
machine learning and its implementation in MLJ. \emph{arXiv:2012.15505
{[}cs{]}}. \url{http://arxiv.org/abs/2012.15505}

@@

\leavevmode\hypertarget{ref-Branco2015SurPre}{}%
Branco, P., Torgo, L., \& Ribeiro, R. (2015). A Survey of Predictive
Modelling under Imbalanced Distributions. \emph{arXiv:1505.01658
{[}cs{]}}. \url{http://arxiv.org/abs/1505.01658}

@@

\leavevmode\hypertarget{ref-Jeni2013FacImb}{}%
Jeni, L. A., Cohn, J. F., \& De La Torre, F. (2013). Facing Imbalanced
DataRecommendations for the Use of Performance Metrics. \emph{2013
Humaine Association Conference on Affective Computing and Intelligent
Interaction}, 245--251. \url{https://doi.org/10.1109/ACII.2013.47}

@@

\leavevmode\hypertarget{ref-Poisot2021ImpMam}{}%
Poisot, T., Ouellet, M.-A., Mollentze, N., Farrell, M. J., Becker, D.
J., Albery, G. F., Gibb, R. J., Seifert, S. N., \& Carlson, C. J.
(2021). Imputing the mammalian virome with linear filtering and singular
value decomposition. \emph{arXiv:2105.14973 {[}q-Bio{]}}.
\url{http://arxiv.org/abs/2105.14973}

@@

\leavevmode\hypertarget{ref-Somodi2017PreDep}{}%
Somodi, I., Lepesi, N., \& Botta-Dukát, Z. (2017). Prevalence dependence
in model goodness measures with special emphasis on true skill
statistics. \emph{Ecology and Evolution}, \emph{7}(3), 863--872.
\url{https://doi.org/10.1002/ece3.2654}

\leavevmode\hypertarget{ref-Steen2021SpaThi}{}%
Steen, V. A., Tingley, M. W., Paton, P. W. C., \& Elphick, C. S. (2021).
Spatial thinning and class balancing: Key choices lead to variation in
the performance of species distribution models with citizen science
data. \emph{Methods in Ecology and Evolution}, \emph{12}(2), 216--226.
\url{https://doi.org/10.1111/2041-210X.13525}
